---
title: "STAT480 Homework2"
author: "Chenz Zhang, NetID chenziz2"
date: "2/6/2019"
output: 
    pdf_document:
      toc: yes
    html_document:
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

```{r}
library(biganalytics)
setwd("~/Stat480/RDataScience/AirlineDelays")
x <- attach.big.matrix("air0708.desc")  #We got this desc from homework1
JulyInds <- which(x[,"Month"] == 7)
Probs <- seq(0.50,0.99,0.01)
delayQuantiles <- quantile(x[JulyInds, "DepDelay"], Probs, 
                           na.rm = TRUE)
print(delayQuantiles)
```

Magnitudes and frequency of delayed departures:  

* The median of delay time is 0 and over half of the flights didn't delay.  (frequency)
* 23% of the fights had serious delay which means they delayed over 15 mins.  (magnitudes)


# Exercise 2  

```{r}
y <- x[JulyInds,]
YearInds <- split(1:length(JulyInds),y[,"Year"])

delayQuantiles2 <- foreach( year = YearInds, .combine=cbind) %do% {
  quantile(y[year, "DepDelay"], Probs, na.rm = TRUE)
}

print(delayQuantiles2)
```

Virtualize data from 2007 to 2008:  
```{r}
library(ggplot2)
df <- data.frame(delay07 = delayQuantiles2[,1], delay08 = delayQuantiles2[,2],
                 quantile = seq(0.50,0.99,0.01))
plot1 <- ggplot(df,aes(quantile)) +
         geom_line(aes(y = delay07, color = "delay07")) + 
         geom_line(aes(y = delay08, color = "delay08")) +
         ylab("Delay Time (Mins)")
plot1
```

* **Difference**: 
    - Delay time in 2007 has larger magnitudes than delay time in 2008 when they are under the same quantile.  
    - 25% of the fights in 2007 and 20% of the fights in 2008 had serious delay which means they delayed over 15 mins.  
    - The median of delay time in 2007 is 0, but the median of delay time in 2008 is -1.  
* **Similarity**: 
    - Delay time in 2007 and 2008 have the same trend (increasing) when quantile increases.  
    - About half of the fights in 2007 and 2008 delayed.  

# Exercise 3  

```{r}
fit1 <- biglm.big.matrix(DepDelay ~ Month + DayOfWeek, data = x)
sumfit1 <- summary(fit1)
summary(fit1)$rsq
sumfit1
```

The model suggests the relationship between DepDelay and the DayOfWeek and Month should be:  
\[DepDelay = 11.2030 - 0.1914 \times Month + 0.1884 \times DayOfWeek\]  

From the r-square result which is 0.0004415451 (extremely small), we can see the model lacks goodness of fit.  
Two predictors are individually significant according to small p-value.  
Month has negative effect on delay time value.(positive effect on delay condition)  
DayOfWeek has positive effect on delay time value.(negative effect on delay condition)  

* Delay time could be influenced by a lot of extra factors sach as location rather than date.  
* Lack of goodness of fit means that Month and DayOfWeek cannot explain delay time by a linear model.  
* Straight linear trend of Month and DayOfWeek cannot meet the real condition.  

Some issues will appear by using this straight linear model:  

* When Month = DayOfWeek = 0, delay time still exists as 11.2030.  
* When the Month increases and DayOfWeek doesn't change, delay time will decrease.  
  This is different from the reality. And it cannot explain seasonal weather change's influence.  


# Exercise 4  

```{r}
fit2 <- biglm.big.matrix(DepDelay ~ I((Month-6)^2) + I(DayOfWeek^2), data = x)
sumfit2 <- summary(fit2)
summary(fit2)$rsq
sumfit2
```

Interpret:  

* Intercept: When Month = 6 and DayOfWeek = 0, delay time is 9.8811.  
* Delays become worse in winter and better in summer. Comparing delay time on the same DayOfWeek, delay in June is the lowest.   
* Quadratic DayOfWeek's coefficient is 0.0235. Delays on weekends is much worse than those on weekdays.  

From the r-square result which is 0.0002049134 (extremely small), we can see the model lacks goodness of fit.  
Two quadratic form predictors are individually significant according to small p-value.  
I((Month - 6)^2) and I(DayOfWeek^2) both have positive effect on delay time value.(negative effect on delay condition)  

This model is even worse than the model in Exercise 3 for smaller r-square. It might because the model in this form is still
uncorrect.  

Therefore, sometimes the model which seems match the real situation performs worse than the original usless model.  


