---
title: "STAT480 Homework3"
author: "Chenz Zhang, NetID chenziz2"
date: "2/9/2019"
output: 
    pdf_document:
      toc: yes
    html_document:
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using code from chapter 3.3, get the sample emails.  

```{r}
spamPath = "~/Stat480/RDataScience/SpamAssassinMessages"

dirNames = list.files(path = paste(spamPath, "messages", 
                                   sep = .Platform$file.sep))
fullDirNames = paste(spamPath, "messages", dirNames, 
                     sep = .Platform$file.sep)

indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
fn = list.files(fullDirNames[1], full.names = TRUE)[indx]
sampleEmail = sapply(fn, readLines)
```


#Exercise 1  

Get hearlist:  
```{r}
splitHeader = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  return(header)
}

sampleHeaderSplit = lapply(sampleEmail, splitHeader)

```

Define my **getSubject** function:  
```{r}
getSubject = function(header) {
  subjectIdx = grep("^ *[Ss]ubject:", header, ignore.case = TRUE)
  if(length(subjectIdx) == 0) return(NA)
  gsub("^ *[Ss]ubject: *(.*)", "\\1", header[subjectIdx], ignore.case = TRUE)
}

```

Demonstrate my function on the headers from 15 samples:  
```{r}
test1 <- lapply(sampleHeaderSplit,getSubject)
unname(test1)
```


#Exercise 2  

Define my **countLetters** function:  
```{r}
countLetters = function(header){
  cleanLetters = tolower(gsub("[[:punct:]0-9[:space:][:blank:]]+", "", getSubject(header)))
  nchar(cleanLetters)
}
```

Demonstrate my function on the headers from 15 samples: 
```{r}
test2 <- lapply(sampleHeaderSplit,countLetters)
unname(test2)
```


#Exercise 3  

Use the **splitMessage** function from chapter 3.5.1 and get split email messages:    
```{r}
splitMessage = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  body = msg[ -(1:splitPoint) ]
  return(list(header = header, body = body))
}

sampleSplit = lapply(sampleEmail, splitMessage)
```

Define **getBoundary** and **dropAttach** functions (from chapter 3.5.2).  
But here, in case of no boundary string in the header, I add one line.  
```{r}
getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  if(length(boundaryIdx) == 0) return(0) # In case of no boundary string
    boundary = gsub('"', "", header[boundaryIdx])
    gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}

dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  if (length(bStringLocs) <= 1) return(body)
  
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )

  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}

```

Define my **countDigits** function:  
```{r}
countDigits = function(message){
  
  msgBoundary = getBoundary(message$header)
  if(msgBoundary == 0) countpart = message$body
  countpart = dropAttach(message$body,msgBoundary)
  countpart = gsub("\\D"," ",countpart)
  
  digits = unlist(strsplit(countpart,"[[:blank:]]"))
  digits = paste(digits,sep = "",collapse = "")
  nchar(digits)
}
```

Here are my results for 15 sample emails:  
```{r}
results = rep(0,15)
for(i in 1:15){
  results[i] = countDigits(sampleSplit[[i]])
}
results
```

#Exercise 4  

```{r}
DirDigits = function(dirN){
  # read all files in the directory
  fileNames = list.files(dirN, full.names = TRUE)
  # drop files that are not email, i.e., cmds
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
  
  messages = lapply(fileNames, readLines, encoding = "latin1")
  # split header and body
  emailSplit = lapply(messages, splitMessage)
  # put body and header in own lists
  bodyList = lapply(emailSplit, function(msg) msg$body)
  headerList = lapply(emailSplit, function(msg) msg$header)
  rm(emailSplit)
  
  # determine which messages have attachments
  hasAttach = sapply(headerList, function(header) {
    CTloc = grep("Content-Type", header)
    if (length(CTloc) == 0) return(0)
    multi = grep("multi", tolower(header[CTloc])) 
    if (length(multi) == 0) return(0)
    multi
  })
  
  hasAttach = which(hasAttach > 0)
  
  # find boundary strings for messages with attachments
  boundaries = sapply(headerList[hasAttach], getBoundary)
  
  # drop attachments from message body
  bodyList[hasAttach] = mapply(dropAttach, bodyList[hasAttach], 
                               boundaries, SIMPLIFY = FALSE)
  
  # new count digits function
  countDigits = function(countpart){
    countpart = gsub("\\D"," ",countpart)
    
    digits = unlist(strsplit(countpart,"[[:blank:]]"))
    digits = paste(digits,sep = "",collapse = "")
    nchar(digits)
  }
  
  dirDigits = unlist(lapply(bodyList, countDigits))
  
  
  
  
  return(dirDigits)
}

```

```{r}
spamPath = "~/Stat480/RDataScience/SpamAssassinMessages"
dirNames = list.files(path = paste(spamPath, "messages", 
                                   sep = .Platform$file.sep))
fullDirNames = paste(spamPath, "messages", dirNames, 
                     sep = .Platform$file.sep)

countDirDigits = sapply(fullDirNames,DirDigits)
```


```{r}
lapply(countDirDigits,summary)
```

According to mean and median, we can divided messages in these 5 directory into 3 groups:  

*  easy_ham (dir: easy_ham and easy_ham_2): small median (<=20) and mean (around 31). So, easy-harm tends to have less digits.  

*  hard_ham (dir: hard_ham): largest median 584.00 and mean 708.95. So, when we roughly seperate ham and spam by one standard for digits, hard_ham trends to be classified as spam. That migtht be why it is hard.  

*  spam (dir: spam and spam_2):  
    +  median (around 60) is larger than easy_ham but much smaller than hard_ham.  
    +  mean (anoud 200) is larger than easy_ham but much smaller than hard_ham.  
    +  maximum is much larger than ham's. We can say spam tends to have larger 'variance'.  

Therefore, when we classify spam and ham emails, range of number of digits might be a good choice.  
